---
title: "Untitled"
format: html
editor: visual
---

```{r}

library(tidyverse)
library(glm2)  
library(grid)
library(vcd)
library(corrplot)    
library(car)         
library(MASS)        
library(stats)      
library(dplyr)
library(tidyr)
library(tidyverse)
library(moderndive)
library(gapminder)
library(sjPlot)
library(stats)
library(jtools)
library(car)
library(nnet)
library(ggplot2)
df <- read_csv("https://raw.githubusercontent.com/RuoningHou/DAS-Group-19/refs/heads/main/dataset19-old.csv")
```

# Check the data structure

```{r}
str(df)
summary(df)
head(df)
```

\` \# Check for missing values

```{r}
colSums(is.na(df))
```

# Histogram: Distribution of the target variable Y

# Histogram of target variable (update 'time_at_shelter' if needed)

```{r}
ggplot(df, aes(x = time_at_shelter)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  theme_minimal() +
  ggtitle("Distribution of Time at Shelter")
```

## Right-Skewed Distribution: The chart shows that the majority of the data is concentrated towards the left side, near zero, and as the shelter stay time increases, the frequency drops rapidly. This is typical of a right-skewed (positively skewed) distribution, indicating that most individuals spent very little time at the shelter, with only a few individuals staying for a longer duration.

# Load visualization package

# Replace 'category1' and 'category2' with actual column names

```{r}
mosaic(~ animal_type + outcome_type, data = df, shade = TRUE, legend = TRUE)
mosaic(~ intake_type + chip_status, data = df, shade = TRUE, legend = TRUE)
mosaic(~ outcome_type + month, data = df, shade = TRUE, legend = TRUE)
mosaic(~ chip_status + year, data = df, shade = TRUE, legend = TRUE)
```

# Replace 'target_variable' and 'category_variable' with actual names

```{r}
ggplot(df, aes(x = outcome_type, y = time_at_shelter)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  ggtitle("Boxplot of Target Variable by Category")
```

##This chart shows the time animals stay in the shelter under different outcome categories (Adoption, Died, Euthanized, Foster, Returned to Owner).

**ADOPTION**:Most animals in this group stay in the shelter for a longer period. The upper range of the boxplot and outliers suggest some animals stay in the shelter for a long time before being adopted.

The median is slightly higher, indicating that animals generally take a longer time to find a new home.

**DIED**:The animals in this group have a shorter stay, with the median close to 0, indicating that these animals were either not adopted or passed away early.

There are no noticeable outliers, suggesting that the time spent by the animals in the shelter is relatively consistent.

**EUTHANIZED**:The animals in this group tend to stay for a longer period, and there are several outliers, suggesting some animals were in the shelter for an extended time before being euthanized.

**FOSTER**:Animals in foster care have a shorter stay, with the median close to 0 and no noticeable outliers, meaning most fostered animals spend a short time in the shelter.

**RETURNED TO OWNER**:Animals returned to their owners also have a short stay, similar to the foster group, with no significant outliers, indicating these animals usually do not stay long in the shelter before being returned.

#Install and load necessary libraries

#Example to pivot the data to long format

```{r}
df_long <- df %>%
  pivot_longer(cols = c(animal_type, intake_type, outcome_type, chip_status),
               names_to = "Category_Type", values_to = "Category")
```

# View the transformed data

```{r}
head(df_long)
```

#group bar charts

# Example plot

```{r}
ggplot(df, aes(x = outcome_type, y = time_at_shelter)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  ggtitle("Boxplot of Target Variable by Category")
```
##This boxplot shows the time animals stay in the shelter under different outcome categories (Adoption, Died, Euthanized, Foster, Returned to Owner).
  ADOPTION:Most animals in this group stay in the shelter for a long period. The upper range of the boxplot and outliers suggest that some animals stayed in the shelter for a long time before being adopted.
  The median is relatively high, indicating that these animals generally take longer to find a new home.
  DIED:The animals in this group have a shorter stay, with the median close to 0, indicating that these animals were either not adopted or passed away early.
  There are no noticeable outliers, suggesting that the time spent by these animals in the shelter is relatively consistent.
  EUTHANIZED:The animals in this group tend to stay in the shelter for a longer period, and there are several outliers, indicating that some animals stayed in the shelter for an extended time before being euthanized.
  FOSTER:Animals in foster care stay in the shelter for a short period. The median is close to 0, and there are no noticeable outliers, indicating that most fostered animals stay briefly in the shelter.
  RETURNED TO OWNER:Animals returned to their owners also have a short stay, similar to the foster group, indicating that these animals are typically returned to their owners after a short stay in the shelter.
# Convert data to long format

```{r}
df_long <- df %>%
  pivot_longer(cols = c(animal_type, intake_type, outcome_type, chip_status),
               names_to = "Category_Type", values_to = "Category")
```

```{r}
ggplot(df_long, aes(x = Category_Type, fill = Category)) +
  geom_bar(position = "fill") +  # Stacked bar plot (relative proportions)
  theme_minimal() +
  ggtitle("Proportional Comparison of 6 Categorical Variables") +
  xlab("Variable") + ylab("Proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
##This chart shows the proportional distribution of four variables: animal type, chip status, intake type, and outcome type:
Animal Type: Dogs have the highest proportion, followed by cats.
Chip Status: Most animals do not have a chip.
Intake Type: Stray animals make up a significant portion.
Outcome Type: Adoption and euthanasia are the most common outcomes, with adoption being the highest.
Overall, dogs and strays dominate, and adoption is the most frequent outcome.
# log transformation

```{r}
df$log_time_at_shelter <- log(df$time_at_shelter + 1)
```

```{r}
ggplot(df, aes(x = outcome_type, y = log_time_at_shelter)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  ggtitle("Boxplot of Target Variable by Category")
```
##This boxplot shows the log-transformed shelter stay time for animals under different outcome categories (Adoption, Died, Euthanized, Foster, Returned to Owner).
Adoption: Animals have a relatively longer stay, with some large outliers.
Died: Animals have a shorter stay, with the median close to 0.
Euthanized: Animals tend to stay longer, with a wider distribution.
Foster: Animals have a shorter stay, with a concentrated distribution.
Returned to Owner: Animals also have a shorter stay, similar to Foster.
Overall, animals in Adoption and Euthanized categories tend to stay longer, while animals in Died, Foster, and Returned to Owner categories have shorter stays.
```{r}
write.csv(df, "Group19_log_transformed.csv", row.names = FALSE)
print("Log-transformed dataset has been saved as 'Group19_log_transformed.csv'")
```

#1.load packages

#2.load data

```{r}
data <- read.csv("https://raw.githubusercontent.com/RuoningHou/DAS-Group-19/refs/heads/main/dataset19.csv")
```

#analysis data

```{r}
evals.data = data
print(evals.data)
ggplot(data = evals.data, aes(x = animal_type, y = time_at_shelter, fill = animal_type)) +
  geom_boxplot() +
  labs(x = "Animal type", y = "Time at shelter")+
  theme(legend.position = "none")
```

#3. GLM analysis

#3.1 GLM analysis of type

#3.1.1 Animal type

```{r}
model_glm.animal_type <- glm(time_at_shelter ~ animal_type, data = data, family = poisson())
summary(model_glm.animal_type)
plot(model_glm.animal_type)
```

#3.1.2 Intake type

```{r}
model_glm.intake_type <- glm(time_at_shelter ~ intake_type, data = data, family = poisson())
summary(model_glm.intake_type)
plot(model_glm.intake_type)
```

#3.1.3 Outcome type

```{r}
model_glm.outcome_type <- glm(time_at_shelter ~ outcome_type, data = data, family = poisson())
summary(model_glm.outcome_type)
plot(model_glm.outcome_type)
```

#3.1.4 Chip status

```{r}
model_glm.chip_status <- glm(time_at_shelter ~ chip_status, data = data, family = poisson())
summary(model_glm.chip_status)
plot(model_glm.chip_status)
```

#3.2 Full Glm analysis

To analyze the impact of different factors on the length of time animals stay in shelters, this paper constructed a generalized linear model (GLM) with time at shelter as the dependent variable and animal type, intake type, outcome type, and chip status as independent variables. $$
  \log \left( E[\text{time_at_shelter}] \right) = \beta_0 + \beta_1 \cdot \text{animal_type} + \beta_2 \cdot \text{intake_type} + \beta_3 \cdot \text{outcome_type} + \beta_4 \cdot \text{chip_status}
$$

```{r}
model.full <- glm(time_at_shelter ~ animal_type + intake_type + outcome_type + chip_status,
                  data = data,
                  family = poisson())
summary(model.full)
plot(model.full)
```

$$
  \begin{aligned}
\log \left( E[\text{time_at_shelter}] \right) =& \\
3.20530 &+ 0.16064 \cdot I(\text{animal_type} = \text{DOG}) \\
&- 0.34465 \cdot I(\text{animal_type} = \text{OTHER}) \\
&- 1.47614 \cdot I(\text{intake_type} = \text{OWNER SURRENDER}) \\
&- 1.09412 \cdot I(\text{intake_type} = \text{STRAY}) \\
&- 0.65638 \cdot I(\text{outcome_type} = \text{EUTHANIZED}) \\
&- 0.54665 \cdot I(\text{outcome_type} = \text{OTHER}) \\
&- 1.47724 \cdot I(\text{outcome_type} = \text{RETURNED TO OWNER}) \\
&+ 0.18684 \cdot I(\text{chip_status} = \text{SCAN CHIP}) \\
&- 0.01124 \cdot I(\text{chip_status} = \text{SCAN NO CHIP}) \\
\end{aligned}
$$

#Final Resolution

```{r}
deviance <- deviance(model.full)
df_residual <- df.residual(model.full)
dispersion <- deviance / df_residual
dispersion
```

dispersion \> 1, indicating overdispersion

#Dealing with overdispersion

```{r}
model.quasi <- glm(time_at_shelter ~ animal_type + intake_type + outcome_type + chip_status,
                   data = data, 
                   family = quasipoisson())

summary(model.quasi)
```

#predicted

```{r}
data$predicted.time <- predict(model.full, type = "response")
print(data$predicted.time)
head(data[, c("time_at_shelter", "predicted.time")])
```

```{r}
plot(data$time_at_shelter, data$predicted.time,
     xlab = "Observed", ylab = "Predicted",
     main = "Observed vs Predicted Time")
abline(0, 1, col = "red", lwd = 2)
```

#VIF

```{r}
vif(model.full)
```

#3.3 AIC analysis

```{r}
AIC(model_glm.animal_type, model_glm.intake_type, model_glm.outcome_type, model_glm.chip_status, model.full)
```

#3.4 BIC analysis

```{r}
BIC(model_glm.animal_type, model_glm.intake_type, model_glm.outcome_type, model_glm.chip_status, model.full)
```
