---
title: "Untitled"
format: html
editor: visual
---

```{r}

library(tidyverse)
library(glm2)  
library(grid)
library(vcd)
library(corrplot)    
library(car)         
library(MASS)        
library(stats)      
library(dplyr)
library(tidyr)
library(tidyverse)
library(moderndive)
library(gapminder)
library(sjPlot)
library(stats)
library(jtools)
library(car)
library(nnet)
library(ggplot2)
df <- read_csv("https://raw.githubusercontent.com/RuoningHou/DAS-Group-19/refs/heads/main/dataset19-old.csv")
```

# Check the data structure

```{r}
str(df)
summary(df)
head(df)
```

# Check for missing values

```{r}
colSums(is.na(df))
```

# Histogram: Distribution of the target variable Y

# Histogram of target variable (update 'time_at_shelter' if needed)

```{r}
ggplot(df, aes(x = time_at_shelter)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  theme_minimal() +
  ggtitle("Distribution of Time at Shelter")
```

#Right-Skewed Distribution: The chart shows that the majority of the data is concentrated towards the left side, near zero, and as the shelter stay time increases, the frequency drops rapidly. This is typical of a right-skewed (positively skewed) distribution, indicating that most individuals spent very little time at the shelter, with only a few individuals staying for a longer duration.

# Load visualization package and replace 'category1' and 'category2' with actual column names

```{r}
mosaic(~ animal_type + outcome_type, data = df, shade = TRUE, legend = TRUE)
mosaic(~ intake_type + chip_status, data = df, shade = TRUE, legend = TRUE)
mosaic(~ outcome_type + month, data = df, shade = TRUE, legend = TRUE)
mosaic(~ chip_status + year, data = df, shade = TRUE, legend = TRUE)
```

# Replace 'target_variable' and 'category_variable' with actual names

```{r}
ggplot(df, aes(x = outcome_type, y = time_at_shelter)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  ggtitle("Boxplot of Target Variable by Category")
```

#This chart shows the time animals stay in the shelter under different outcome categories (Adoption, Died, Euthanized, Foster, Returned to Owner).

**ADOPTION**:Most animals in this group stay in the shelter for a longer period. The upper range of the boxplot and outliers suggest some animals stay in the shelter for a long time before being adopted.

The median is slightly higher, indicating that animals generally take a longer time to find a new home.

**DIED**:The animals in this group have a shorter stay, with the median close to 0, indicating that these animals were either not adopted or passed away early.

There are no noticeable outliers, suggesting that the time spent by the animals in the shelter is relatively consistent.

**EUTHANIZED**:The animals in this group tend to stay for a longer period, and there are several outliers, suggesting some animals were in the shelter for an extended time before being euthanized.

**FOSTER**:Animals in foster care have a shorter stay, with the median close to 0 and no noticeable outliers, meaning most fostered animals spend a short time in the shelter.

**RETURNED TO OWNER**:Animals returned to their owners also have a short stay, similar to the foster group, with no significant outliers, indicating these animals usually do not stay long in the shelter before being returned.

# Install and load necessary libraries

# Example to pivot the data to long format

```{r}
df_long <- df %>%
  pivot_longer(cols = c(animal_type, intake_type, outcome_type, chip_status),
               names_to = "Category_Type", values_to = "Category")
```

# View the transformed data

```{r}
head(df_long)
```

#group bar charts

# Example plot

```{r}
ggplot(df, aes(x = outcome_type, y = time_at_shelter)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  ggtitle("Boxplot of Target Variable by Category")
```

#This boxplot shows the time animals stay in the shelter under different outcome categories (Adoption, Died, Euthanized, Foster, Returned to Owner). ADOPTION:Most animals in this group stay in the shelter for a long period. The upper range of the boxplot and outliers suggest that some animals stayed in the shelter for a long time before being adopted. The median is relatively high, indicating that these animals generally take longer to find a new home. DIED:The animals in this group have a shorter stay, with the median close to 0, indicating that these animals were either not adopted or passed away early. There are no noticeable outliers, suggesting that the time spent by these animals in the shelter is relatively consistent. EUTHANIZED:The animals in this group tend to stay in the shelter for a longer period, and there are several outliers, indicating that some animals stayed in the shelter for an extended time before being euthanized. FOSTER:Animals in foster care stay in the shelter for a short period. The median is close to 0, and there are no noticeable outliers, indicating that most fostered animals stay briefly in the shelter. RETURNED TO OWNER:Animals returned to their owners also have a short stay, similar to the foster group, indicating that these animals are typically returned to their owners after a short stay in the shelter.

# Convert data to long format

```{r}
df_long <- df %>%
  pivot_longer(cols = c(animal_type, intake_type, outcome_type, chip_status),
               names_to = "Category_Type", values_to = "Category")
```

```{r}
ggplot(df_long, aes(x = Category_Type, fill = Category)) +
  geom_bar(position = "fill") +  # Stacked bar plot (relative proportions)
  theme_minimal() +
  ggtitle("Proportional Comparison of 6 Categorical Variables") +
  xlab("Variable") + ylab("Proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#This chart shows the proportional distribution of four variables: animal type, chip status, intake type, and outcome type: Animal Type: Dogs have the highest proportion, followed by cats. Chip Status: Most animals do not have a chip. Intake Type: Stray animals make up a significant portion. Outcome Type: Adoption and euthanasia are the most common outcomes, with adoption being the highest. Overall, dogs and strays dominate, and adoption is the most frequent outcome. \# log transformation

```{r}
df$log_time_at_shelter <- log(df$time_at_shelter + 1)
```

```{r}
ggplot(df, aes(x = outcome_type, y = log_time_at_shelter)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  ggtitle("Boxplot of Target Variable by Category")
```

#This boxplot shows the log-transformed shelter stay time for animals under different outcome categories (Adoption, Died, Euthanized, Foster, Returned to Owner). Adoption: Animals have a relatively longer stay, with some large outliers. Died: Animals have a shorter stay, with the median close to 0. Euthanized: Animals tend to stay longer, with a wider distribution. Foster: Animals have a shorter stay, with a concentrated distribution. Returned to Owner: Animals also have a shorter stay, similar to Foster. Overall, animals in Adoption and Euthanized categories tend to stay longer, while animals in Died, Foster, and Returned to Owner categories have shorter stays.

```{r}
write.csv(df, "Group19_log_transformed.csv", row.names = FALSE)
print("Log-transformed dataset has been saved as 'Group19_log_transformed.csv'")
```

# 1.load packages

# 2.load data

```{r}
data <- read.csv("https://raw.githubusercontent.com/RuoningHou/DAS-Group-19/refs/heads/main/dataset19.csv")
```

#analysis data

```{r}
evals.data = data
print(evals.data)
ggplot(data = evals.data, aes(x = animal_type, y = time_at_shelter, fill = animal_type)) +
  geom_boxplot() +
  labs(x = "Animal type", y = "Time at shelter")+
  theme(legend.position = "none")
```

#This box plot shows the shelter stay duration (time_at_shelter) for different animal types (CAT, DOG, OTHER).

# 2.1. Overall Trends:

Median stay durations are similar across all categories, but DOG has a wider distribution.

CAT and DOG show greater variability, while OTHER has a smaller range.

Outliers (black dots) are more frequent in DOG and CAT, indicating some animals stay significantly longer.

# 2.2. Key Observations:

Dogs tend to stay longer than cats, as seen in their higher IQR and max values.

Other animals leave shelters faster, possibly due to smaller sample sizes or different handling processes.

All distributions are right-skewed, meaning most animals have short stays, with a few staying much longer.

# 2.3. Possible Causes:

Dogs require longer adoption matching than cats.

Other animals (e.g., small mammals, reptiles) might have faster processing.

Factors like microchip status, health, and age could further influence shelter stay durations.

# 3. GLM analysis

# 3.1 GLM analysis of type

# 3.1.1 Animal type

```{r}
model_glm.animal_type <- glm(time_at_shelter ~ animal_type, data = data, family = poisson())
summary(model_glm.animal_type)
plot(model_glm.animal_type)
```

#### 1. Model Interpretation

This Poisson regression model examines the effect of animal type (CAT, DOG, OTHER) on shelter stay duration (time_at_shelter). Key findings:

Intercept = 1.6918, representing the Poisson coefficient for the reference category (CAT).

DOG coefficient = 0.1533, p \< 0.001, indicating that dogs stay significantly longer than cats.

OTHER coefficient = 0.0069, p = 0.951, showing no significant difference from cats.

AIC = 18257, suggesting the model may not fully explain shelter stay duration.

#### 2. Residual Diagnostics

Residuals vs. Fitted Plot: Large outliers indicate possible overdispersion.

Q-Q Plot: Deviations at the upper end suggest a poor Poisson fit.

Scale-Location Plot: Variance is stable, but extreme values exist.

Residuals vs. Leverage Plot: Some points have high influence (Cook’s distance).

#### 3. Conclusion & Improvements

Dogs stay significantly longer than cats; OTHER animals show no difference.

# 3.1.2 Intake type

```{r}
model_glm.intake_type <- glm(time_at_shelter ~ intake_type, data = data, family = poisson())
summary(model_glm.intake_type)
plot(model_glm.intake_type)
```

#### 1. Model Interpretation

This Poisson regression examines the effect of intake type on shelter stay duration:

Intercept = 2.5294, representing the reference category .

OWNER SURRENDER: Coefficient = -0.8802, p \< 0.001, indicating significantly shorter stays.

STRAY: Coefficient = -0.7293, p \< 0.001, also showing shorter stays.

AIC = 17705, better fit than the previous model (AIC = 18257).

#### 2. Residual Diagnostics

Residuals vs. Fitted Plot: High outliers suggest overdispersion.

Q-Q Plot: Deviations indicate Poisson distribution mismatch.

Scale-Location Plot: Stable variance but extreme values.

Residuals vs. Leverage: Some high-influence points.

#### 3. Conclusion & Suggestions

Owner-surrendered and stray animals have significantly shorter stays.

# 3.1.3 Outcome type

```{r}
model_glm.outcome_type <- glm(time_at_shelter ~ outcome_type, data = data, family = poisson())
summary(model_glm.outcome_type)
plot(model_glm.outcome_type)
```

#### 1. Model Interpretation

This Poisson regression examines the effect of outcome type on shelter stay duration:

Intercept = 2.1594, representing the adopted category.

EUTHANIZED: Coefficient = -0.5849, p \< 0.001, indicating shorter stays.

OTHER: Coefficient = -0.6093, p \< 0.001, also showing shorter stays.

RETURNED TO OWNER: Coefficient = -1.0617, p \< 0.001, with the shortest stay duration.

AIC = 16667, suggesting better model fit than previous (AIC = 17705).

#### 2. Residual Diagnostics

High outliers indicate overdispersion.

Q-Q plot shows deviation, suggesting Poisson may not fully fit.

Stable variance but extreme values in the scale-location plot.

Influential points detected in residuals vs. leverage plot.

#### 3. Conclusion

Animals returned to owners have the shortest stay, while euthanized and other outcomes also have shorter durations.

# 3.1.4 Chip status

```{r}
model_glm.chip_status <- glm(time_at_shelter ~ chip_status, data = data, family = poisson())
summary(model_glm.chip_status)
plot(model_glm.chip_status)
```

#### 1. Model Interpretation

This Poisson regression examines chip status and its effect on shelter stay duration:

Intercept = 1.6933, representing the “not scanned” category.

SCAN CHIP: Coefficient = 0.1161, p = 0.0286, indicating slightly longer stays for animals with chips.

SCAN NO CHIP: Coefficient = 0.1275, p = 0.0094, showing a similar effect for animals without chips.

AIC = 18293, suggesting moderate model fit.

#### 2. Residual Diagnostics

High outliers suggest overdispersion.

Q-Q plot shows deviation, indicating a poor Poisson fit.

Stable variance but extreme values in the scale-location plot.

Influential points detected in the leverage plot.

#### 3. Conclusion

Animals with or without chips have slightly longer stays, likely due to chip registration and adoption processes.

# 3.2 Full Glm analysis

To analyze the impact of different factors on the length of time animals stay in shelters, this paper constructed a generalized linear model (GLM) with time at shelter as the dependent variable and animal type, intake type, outcome type, and chip status as independent variables. $$
  \log \left( E[\text{time_at_shelter}] \right) = \beta_0 + \beta_1 \cdot \text{animal_type} + \beta_2 \cdot \text{intake_type} + \beta_3 \cdot \text{outcome_type} + \beta_4 \cdot \text{chip_status}
$$

```{r}
model.full <- glm(time_at_shelter ~ animal_type + intake_type + outcome_type + chip_status,
                  data = data,
                  family = poisson())
summary(model.full)
plot(model.full)
```

$$
  \begin{aligned}
\log \left( E[\text{time_at_shelter}] \right) =& \\
3.20530 &+ 0.16064 \cdot I(\text{animal_type} = \text{DOG}) \\
&- 0.34465 \cdot I(\text{animal_type} = \text{OTHER}) \\
&- 1.47614 \cdot I(\text{intake_type} = \text{OWNER SURRENDER}) \\
&- 1.09412 \cdot I(\text{intake_type} = \text{STRAY}) \\
&- 0.65638 \cdot I(\text{outcome_type} = \text{EUTHANIZED}) \\
&- 0.54665 \cdot I(\text{outcome_type} = \text{OTHER}) \\
&- 1.47724 \cdot I(\text{outcome_type} = \text{RETURNED TO OWNER}) \\
&+ 0.18684 \cdot I(\text{chip_status} = \text{SCAN CHIP}) \\
&- 0.01124 \cdot I(\text{chip_status} = \text{SCAN NO CHIP}) \\
\end{aligned}
$$

#### 1. Model Interpretation

This Poisson model analyzes shelter stay duration based on animal type, intake type, outcome type, and chip status:

DOG (0.1606, p \< 0.001) → Longer stay than reference.

OWNER SURRENDER (-1.4761, p \< 0.001) & STRAY (-1.0941, p \< 0.001) → Shorter stays.

EUTHANIZED (-0.5663, p \< 0.001) & RETURNED TO OWNER (-1.0742, p \< 0.001) → Shortest stays.

SCAN CHIP (0.1868, p \< 0.001) → Slightly longer stay.

AIC = 15205, improved fit over single-variable models.

#### 2. Residual Analysis

Scattered residuals, some outliers present.

Q-Q plot shows deviation from normality.

Stable variance, high Cook’s distance for some points.

#### 3. Conclusion

Dogs stay longer; surrendered & returned animals have the shortest stays. Chipped animals stay slightly longer, likely due to adoption processing.

#Final Resolution

```{r}
deviance <- deviance(model.full)
df_residual <- df.residual(model.full)
dispersion <- deviance / df_residual
dispersion
```

dispersion \> 1, indicating overdispersion

#Dealing with overdispersion

```{r}
model.quasi <- glm(time_at_shelter ~ animal_type + intake_type + outcome_type + chip_status,
                   data = data, 
                   family = quasipoisson())

summary(model.quasi)
```

#### 1. Overdispersion Adjustment

Quasi-Poisson regression (family = quasipoisson) was used to correct overdispersion.

Dispersion parameter = 6.8928, confirming excess variance.

#### 2. Key Findings

DOG (0.1606, p = 0.0149) → Longer shelter stay.

OWNER SURRENDER (-1.4761, p \< 0.001) & STRAY (-1.0941, p \< 0.001) → Shorter stays.

SCAN CHIP (0.1868, p = 0.1925) → Not significant.

#### 3. Conclusion

Quasi-Poisson corrected overdispersion, but high outliers remain.

Negative Binomial Regression may provide a better fit.

Consider interaction terms or additional variables to improve the model.

# predicted

```{r}
data$predicted.time <- predict(model.full, type = "response")
print(data$predicted.time)
head(data[, c("time_at_shelter", "predicted.time")])
```

```{r}
plot(data$time_at_shelter, data$predicted.time,
     xlab = "Observed", ylab = "Predicted",
     main = "Observed vs Predicted Time")
abline(0, 1, col = "red", lwd = 2)
```

#### 1. Overview

Poisson regression (`model.full`) was used to predict shelter stay duration (`predicted.time`).

#### 2. Key Observations

Predictions underestimate actual stay times, especially for long durations (e.g., 21 days predicted as 8.16).

Short stays (0-2 days) are predicted more accurately.

#### 3. Conclusion & Recommendations

Overdispersion or missing variables may cause errors.

Consider Negative Binomial Regression to better handle long-tail data.

# VIF

```{r}
vif(model.full)
```

#### 1. Overview

VIF detects multicollinearity (VIF \> 5 or 10 is concerning).

GVIF is used for categorical variables, with GVIF\^(1/(2\*Df)) for comparison.

#### 2. Key Findings

All GVIF values are near 1 (highest: 1.0559 for `intake_type`).

No significant multicollinearity among variables.

#### 3. Conclusion

Model stability is not affected by collinearity.

Consider checking interactions for potential risks.

# 3.3 AIC analysis

```{r}
AIC(model_glm.animal_type, model_glm.intake_type, model_glm.outcome_type, model_glm.chip_status, model.full)
```

# 3.4 BIC analysis

```{r}
BIC(model_glm.animal_type, model_glm.intake_type, model_glm.outcome_type, model_glm.chip_status, model.full)
```

-   AIC Analysis:

    model.full (AIC = 15205.41) is the best model, significantly lower than single-variable models, indicating improved fit with multiple variables.

-   BIC Analysis:

```         
-   **model.full (BIC = 15260.66) remains optimal**, though slightly higher due to BIC's stronger penalty on model complexity.
```

-   Final Conclusion:

```         
-   **model.full is the best choice**, as it has the lowest AIC and BIC, ensuring a high model fit without excessive complexity.
```
